The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-19 paper.tex
Revisoes do Roberto: - reforcar diferencas a Esterel (em vez de semelhancas) ja no inicio da secao 2 - incluir analise estatica de shared-memory concurrency - reescrever a discussao do exemplo de finalizacao, deixando claro que ele e' "unsafe" no inicio, e que a chamada a enqueue executa em "background"

Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel and 
\CEU for the following control specification~\cite{esterel.primer}:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''.}
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (lines 4--9, in both 
implementations), as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (lines 3--10), which serves the same purpose of \CEU's 
\code{par/or} (lines 3--12):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the \code{loop}.

\begin{figure}[t]
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
loop
   abort
      [
         await A
      $\|$
         await B
      ];
      emit O
   when R
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ A control specification implemented in Esterel and \CEU:
\emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
%\newline
{\small
%TODO.
}
\label{lst.abro}
}
\end{figure}

\CEU, following the Esterel mindset, has a strong imperative flavor, with 
explicit control flow through sequences, loops, parallels, and also 
assignments.
Being designed for control-intensive applications, it provides support for 
concurrent lines of execution and broadcast communication through events.
%
In addition, \CEU also employs the synchronous model, in whicn programs advance 
in a sequence of discrete reactions to external events.
Internal computations within a reaction (e.g. expressions, assignments, and 
native calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
The \code{await} statements are the only ones that halt a running reaction and 
allow a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.

In the sections that follow, we review the main differences between \CEU and 
Esterel~\cite{ceu.sensys13}: deterministic execution 
(Section~\ref{sec.ceu.det}), safe abortion with finalization 
(Section~\ref{sec.ceu.abrt}), first-class timers 
(Section~\ref{sec.ceu.timers}), and stack-based internal events 
(Section~\ref{sec.ceu.ints}).

%\newpage %TTT
\subsection{Deterministic Execution}
\label{sec.ceu.det}

Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
However, the execution order for operations with side-effects within a reaction 
is non-deterministic: ``if there is no control dependency, as in \code{"call 
f1() || call f2()"}, the order is unspecified and it would be an error to rely 
on it''~\cite{esterel.primer}.
%
In \CEU, when multiple trails are active at a time, they are scheduled in the 
order they appear in the program source code.
%
Hence, \CEU is deterministic also with respect to the order of execution of 
side effects within a reaction~\cite{ceu.sensys13}.
%
Figure~\ref{lst.det} compares the two syntactically equivalent code fragments 
in Esterel and \CEU to illustrate the semantic difference regarding 
(non-)determinism.

On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
In contrast, Esterel forbids sharing memory: \emph{``if a variable is written 
by some thread, then it can neither be read nor be written by concurrent 
threads''}~\cite{esterel.primer}.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for I/O and the lack of hardware support for real parallelism.
Other embedded languages made a similar design choice~\cite{wsn.sol,pret}.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

TODO: names (other embedded languages)

\begin{figure}[b]
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
[
    call f1();
$\|$
    call f2();
];
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
par/and do
    _f1();
with
    _f2();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ In Esterel, the execution order between \code{f1} and \code{f2} is 
unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2}.
{\small
%TODO.
}
\label{lst.det}
}
\end{figure}

@@

\begin{figure}[t]
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void STOP, RETRANSMIT, SENDACK;
par/or do
    await STOP;
with
    loop do
        par/or do
            await RETRANSMIT;
        with
            par/and do
                await 1min;
            with
                var _pkt_t buffer;
                <fill-buffer-info>
                _send_enqueue(&buffer);
                await SENDACK;
            end
        end
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.41\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=11]
<...>
    var _pkt_t buffer;
    <fill-buffer-info>
    finalize
        _send_enqueue(&buffer);
    with
        _send_cancel(&buffer);
    end
    await SENDACK;
<...>
\end{lstlisting}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{A network protocol in \CEU extended with a finalization clause to 
cancel a transmission.
%\caption{ Parallel compositions can describe complex state machines.
\label{lst.fin}
}
\end{figure}

To illustrate threats related to abortion, consider the example in the left of 
Figure~\ref{lst.fin} with hierarchical \code{par/or} and \code{par/and} 
compositions describing the state machine of a data collection protocol for 
sensor networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (line 1) 
represent the external interface of the protocol with a client application.
%
The protocol has two trails:
one simply monitors the stopping event (line 3);
the other periodically transmits a packet (lines 5--18).
%
The periodic transmission is a loop that starts two other trails (lines 6--17):
one handles an immediate retransmission request (line 7);
the other transmits the packet%
\footnote{
The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
with external \emph{C} libraries explicit and are required in \CEU.
}
and waits for a confirmation (lines 9--16).
%
%
The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
takes at least one minute before looping, to avoid flooding the network with 
packets.
%
At any time, the client may request a retransmission (line 7), which terminates 
the \code{par/or} (line 6), aborts the ongoing transmission (line 14, if not 
idle), and restarts the loop (line 5).
%
The client may also request to stop the whole protocol at any time (line 3).
%
Note that the \code{buffer} packet (line 12) is a local variable whose address 
is passed to function \code{\_send\_enqueue} for transmission.
The call enqueues the pointer in the radio driver, which holds it up to the 
emission of \code{SENDACK} acknowledging the packet transmission.
%
In the meantime, if the sending trail is be aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

@@

\begin{figure}[t]
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
// ESTEREL
abort
    await A;
    call f();
when A;

@@

\begin{figure}[t]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}
var int v;
await 10ms;
v = 1;
await 1ms;
v = 2;
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.51\linewidth}
\begin{lstlisting}
par/or do
    await 10ms;
    <...>         // any non-awaiting sequence
    await  1ms;
    v = 1;
with
    await 12ms;
    v = 2;
end
\end{lstlisting}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{ First-class timers in \CEU.
\label{lst.timers}
}
\end{figure}

\CEU also takes into account the fact that time is a physical quantity that can 
be added and compared.
For instance, for the program in the right of Figure~\ref{lst.timers}, although 
the scheduler cannot guarantee that the first trail terminates exactly in 11ms, 
it can at least ensure that the program always terminates with \code{v=1}:
%
Given that any non-awaiting sequence is considered to take no time in the 
synchronous model, the first trail is guaranteed to terminate before the second 
trail, because $10+1 < 12$.
%
A similar program in a language without first-class support for timers, would 
depend on the execution timings for the code marked as \code{<...>}, making the 
reasoning about the execution behavior more difficult.

@@

\begin{figure}%[b]
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[mathescape=true]
// ESTEREL
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
// CEU
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU. \newline
\label{lst.prints}
}
\end{figure}

@@

\begin{figure}[h]
\begin{lstlisting}[mathescape=true]
                                   // primary expressions
  p ::= mem(id)                   (any memory access to `id')
      $|$ await(id)                  (await event `id')
      $|$ emit(id)                   (emit event `id')
      $|$ break                      (loop escape)
                                   // compound expressions
      $|$ if mem(id) then p else p   (conditional)
      $|$ p ; p                      (sequence)
      $|$ loop p                     (repetition)
      $|$ p and p                    (par/and)
      $|$ p or p                     (par/or)
      $|$ fin p                      (finalization)
                                   // derived by semantic rules
      $|$ awaiting(id,n)             (awaiting `id' since sequence number `n')
      $|$ emitting(n)                (emitting on stack level `n')
      $|$ p @ loop p                 (unwinded loop)
\end{lstlisting}
\caption{
    Reduced syntax of \CEU.
\label{lst.formal.syntax}
}
\end{figure}

@@

The core of our semantics is a relation that, given a sequence number $n$ 
identifying the current reaction chain, maps a program $p$ and a stack of 
events $S$ in a single step to a modified program and stack:
%
\begin{align*}
\LL S, p \RR &\ST
\LL S', p' \RR
    & \textbf{(relation-inner)}
\end{align*}
%
where
%
\begin{align*}
S, S' &\in id^*
    &&(sequence~of~event~identifiers: [id_{top}, ..., id_{bottom}]) \\
p, p' &\in P
    && (program~as~described~in~Figure~\ref{lst.formal.syntax}) \\
n     &\in \mathds{N}
    && (unique~identifier~for~the~reaction~chain)
\end{align*}

At the beginning of a reaction chain, the stack is initialized with the 
occurring external event $ext$ ($S=[ext]$), but $emit$ expressions can push new 
events on top of it (we discuss how they are popped further).
The ever-increasing sequence number $n$ will prevent that awaiting expressions 
awake in the same reaction they are reached (ensuring that the program can 
progress).

@@

We describe this relation with a set of \emph{small-step} structural semantics 
rules, which are built in such a way that at most one transition is possible at 
any time, resulting in deterministic reaction chains.
%
The transition rules for the primary expressions are as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL S,\1await(id) \RR &\ST
\LL S,\1awaiting(id,n+1) \RR
    & \textbf{(awaiting)}      \\
%%%
\LL id:S,\1awaiting(id,m) \RR &\ST
\LL id:S,\1nop \RR, \2m<=n
    & \textbf{(awake)}   \\
%%%
\LL S,\1emit(id) \RR &\ST
\LL id:S,\1emitting(|S|) \RR
    & \textbf{(emitting)}       \\
%%%
\LL S,\1emitting(|S|) \RR &\ST
\LL S,\1nop \RR
    & \textbf{(pop)}
\end{align*}
}

An $await$ is simply transformed into an $awaiting$ (rule \textbf{awaiting}) as 
an artifact to remember the external sequence number $n$ it can awake.
An $awaiting$ can only transit to a $nop$ (rule \textbf{awake}) if its referred 
event $id$ matches the top of the stack and its sequence number is smaller or 
equal than the current one ($m<n$).
%
%Remember that in \CEU, the \code{await} statement returns the value associated 
%with the corresponding event: the yielded $mem$ represents the operation to 
%query that value.
%
An $emit$ transits to an $emitting$ holding the current stack level ($|S|$ 
stands for the stack length), and pushing the referred event on the stack (rule 
\textbf{emitting}).
With the new stack level $|S|+1$, the $emitting(|S|)$ itself cannot transit, as 
rule \textbf{pop} expects its parameter to match the current stack level.
This trick provides the desired stack-based semantics for internal events.

Proceeding to compound expressions, the rules for conditionals and sequences 
are straightforward:

{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS val(id,n) \neq 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, p \RR }
    & \textbf{(if-true)}       \\
%%%
& \frac
    { \DS val(id,n) = 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, q \RR }
    & \textbf{(if-false)}       \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~;~q) \RR \ST \LL S', (p'~;~q) \RR }
    & \textbf{(seq-adv)}      \\
%%%
& \LL S, (mem(id)~;~q) \RR \ST  \LL S, q \RR
    & \textbf{(seq-nop)}      \\
%%%
& \LL S, (break~;~q) \RR \ST \LL S, break \RR
    & \textbf{(seq-brk)}
\end{eqnarray*}
}

Given that our semantics focuses on control, rules \textbf{if-true} and 
\textbf{if-false} are the only to query $mem$ expressions.
%
The ``magical'' function $val$ receives the memory identifier and current 
reaction sequence number, returning the current memory value.
%
Although the value here is arbitrary, it is unique in a reaction chain, because 
a given expression can execute only once within it (remember that $loops$ must 
contain $awaits$ which, from rule \textbf{awaiting}, cannot awake in the same 
reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

The rules for loops are analogous to sequences, but use \code{`@'} as 
separators to properly bind breaks to their enclosing loops:

{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (loop~p) \RR \ST \LL S, (p~@~loop~p) \RR
    & \textbf{(loop-expd)}       \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
% -----------------------------------------------------------
    {\DS \LL S, (p~@~loop~q) \RR \ST \LL S', (p'~@~loop~q) \RR }
    & \textbf{(loop-adv)}    \\
%%%
& \LL S, (mem(id)~@~loop~p) \RR \ST \LL S, loop~p \RR
    & \textbf{(loop-nop)}    \\
%%%
& \LL S, (break~@~loop~p) \RR \ST \LL S, nop \RR
    & \textbf{(loop-brk)}
\end{eqnarray*}
}

>>>>

In the sections that follow, we review the main differences between \CEU and 
Esterel~\cite{ceu.sensys13}:
shared-memory concurrency (Section~\ref{sec.ceu.det}),
safe abortion with finalization (Section~\ref{sec.ceu.abrt}),
first-class timers (Section~\ref{sec.ceu.timers}),
and stack-based internal events (Section~\ref{sec.ceu.ints}).

Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
implementations in Esterel and \CEU for the following control 
specification~\cite{esterel.primer}:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''.}
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (lines 4--9, in both 
implementations), as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (lines 3--10), which serves the same purpose of \CEU's 
\code{par/or} (lines 3--12):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the \code{loop}.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
loop
   abort
      [
         await A
      $\|$
         await B
      ];
      emit O
   when R
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ A control specification implemented in Esterel and \CEU:
\emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
%\newline
{\small
%TODO.
}
\label{lst.abro}
}
\end{figure}

\CEU, following the Esterel mindset, has a strong imperative flavor, with 
explicit control flow through sequences, loops, parallels, and also 
assignments.
Being designed for control-intensive applications, it provides support for 
concurrent lines of execution and broadcast communication through events.
%
In addition, \CEU also employs the synchronous model, in which programs advance 
in a sequence of discrete reactions to external events.
Internal computations within a reaction (e.g. expressions, assignments, and 
native calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
The \code{await} statements are the only ones that halt a running reaction and 
allow a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.

%\newpage %TTT
\subsection{Shared-Memory Concurrency}
\label{sec.ceu.det}

Embedded applications make extensive use of shared memory, such as for sharing 
resources through memory-mapped registers.
Hence, an important goal of \CEU is to ensure a reliable execution for 
concurrent programs that share memory.
%
Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations with side-effects within a reaction 
is non-deterministic: ``if there is no control dependency, as in \code{"call 
f1() || call f2()"}, the order is unspecified and it would be an error to rely 
on it''~\cite{esterel.primer}.
%
Therefore, Esterel forbids sharing memory: \emph{``if a variable is written by 
some thread, then it can neither be read nor be written by concurrent 
threads''}~\cite{esterel.primer}.

Concurrency in \CEU is characterized when two or more trail segments in 
parallel execute during the same reaction chain.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
In the program in the left of Figure~\ref{lst.shared}, the two assignments to 
\code{x} can never run concurrently, because each trail segment reacts to a 
different input event and, according to the synchronous model, each reaction is 
atomic.
%
However, the program in the right is non-deterministic, because the two 
assignments to \code{y} occur in the same reaction to input \code{A}.

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
the code in the left is safe because the trails access \code{x} atomically in 
different reactions;
the code in the right is unsafe because both trails access \code{y} in the same 
reaction.
%\newline
{\small
%TODO.
}
\label{lst.shared}
}
\end{figure}

\CEU performs a temporal analysis at compile time and detects concurrent 
accesses to shared variables, as follows~\cite{ceu.sensys13}:
\emph{if a variable is written in a trail segment, then a concurrent trail 
segment cannot read or write to that variable, nor dereference a pointer of 
that variable type.}
An analogous policy is applied for pointers \emph{vs} variables and pointers 
\emph{vs} pointers, as well as for system calls with side effects (e.g., 
\code{printf}).
The algorithm for the analysis holds the set of all events in preceding 
\code{await} statements for each variable access.
Then, the sets for all accesses in parallel trails are compared to assert that 
no events are shared among them.
Otherwise the compiler warns about the suspicious accesses.

Regardless of the static analysis, when multiple trails are active at a time in 
\CEU, they are scheduled in the order they appear in the program source code.
%
Therefore, even though the program in the right of Figure~\ref{lst.shared} is 
suspicious, the assignments to \code{y} both atomic and deterministic, i.e., 
after \code{A} occurs, \code{y} is $4$ (\code{(1+1)*2}).
%
On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for I/O and the lack of hardware support for real parallelism.
Other synchronous embedded languages, such as SOL~\cite{wsn.sol} and 
PRET-C\cite{rp.pretc}, made a similar design choice.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

Figure~\ref{lst.det} compares the two syntactically equivalent code fragments 
in Esterel and \CEU to summarize the semantic difference regarding 
(non-)determinism.
%
Even though the program in \CEU executes deterministically, the compiler still 
issues a warning, because an apparently innocuous reordering of trails modifies 
the semantics of the program.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
[
    call f1();
$\|$
    call f2();
];
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
par/and do
    _f1();
with
    _f2();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ In Esterel, the execution order between \code{f1} and \code{f2} is 
unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2}.
{\small
%TODO.
}
\label{lst.det}
}
\end{figure}

@@

\begin{figure}
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void STOP, RETRANSMIT, SENDACK;
par/or do
    await STOP;
with
    loop do
        par/or do
            await RETRANSMIT;
        with
            par/and do
                await 1min;
            with
                var _pkt_t buffer;
                <fill-buffer-info>
                _send_enqueue(&buffer);
                await SENDACK;
            end
        end
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.41\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=11]
<...>
    var _pkt_t buffer;
    <fill-buffer-info>
    finalize
        _send_enqueue(&buffer);
    with
        _send_cancel(&buffer);
    end
    await SENDACK;
<...>
\end{lstlisting}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{The unsafe network protocol in the left is extended with a 
finalization clause in the right to handle abortion properly.
%\caption{ Parallel compositions can describe complex state machines.
\label{lst.fin}
}
\end{figure}

To illustrate threats related to abortion, consider the unsafe example in the 
left of Figure~\ref{lst.fin}, describing the state machine of a data collection 
protocol for sensor networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (line 1) 
represent the external interface of the protocol with a client application.
%
The protocol has to transmit a packet every minute, unless it receives a 
\code{RETRANSMIT} request to immediately re-transmit, or a \code{STOP} request 
to terminate.
%
The protocol is composed of two main trails:
one simply monitors the stopping event (line 3);
the other periodically transmits a packet (lines 5--18).
%
The periodic transmission is a loop that starts two other trails (lines 6--17):
one handles the immediate retransmission request (line 7);
the other transmits the packet%
\footnote{
The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
with external \emph{C} libraries explicit and are required in \CEU.
}
and waits for a confirmation (lines 9--16).
%
%
The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
takes at least one minute before looping, to avoid flooding the network with 
packets.
Note that the call to \code{\_send\_enqueue} is \emph{asynchronous}, handing to 
the radio driver a pointer to the lexically-scoped packet.
The driver will make the transmission in the background, holding the packet 
until it signals the application with the \code{SENDACK} to acknowledge 
completion.
%
At any time, the client may request a retransmission (line 7), which terminates 
the \code{par/or} (line 6), aborts the ongoing transmission (line 14, if not 
idle), and restarts the loop (line 5).
%
The client may also request to stop the whole protocol at any time (line 3).
%
In the meantime, if the sending trail is be aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver, which will possibly transmit 
wrong data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
// ESTEREL
abort
    await A;
    call f();
when A;

@@

\begin{figure}
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}
var int v;
await 10ms;
v = 1;
await 1ms;
v = 2;
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.51\linewidth}
\begin{lstlisting}
par/or do
    await 10ms;
    <...>         // any non-awaiting sequence
    await  1ms;
    v = 1;
with
    await 12ms;
    v = 2;
end
\end{lstlisting}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{ First-class timers in \CEU.
\label{lst.timers}
}
\end{figure}

\CEU also takes into account the fact that time is a physical quantity that can 
be added and compared.
For instance, for the program in the right of Figure~\ref{lst.timers}, although 
the scheduler cannot guarantee that the first trail terminates exactly in 11ms, 
it can at least ensure that the program always terminates with \code{v=1}:
%
Given that any non-awaiting sequence is considered to take no time in the 
synchronous model, the first trail is guaranteed to terminate before the second 
trail, because $10+1 < 12$.
%
A similar program in a language without first-class support for timers would 
depend on the execution timings for the code marked as \code{<...>}, making the 
reasoning about the execution behavior more difficult.

@@

\begin{figure}
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[mathescape=true]
// ESTEREL
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
// CEU
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU. \newline
\label{lst.prints}
}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}[mathescape=true]
                                   // primary expressions
  p ::= mem(id)                   (any memory access to `id')
      $|$ await(id)                  (await event `id')
      $|$ emit(id)                   (emit event `id')
      $|$ break                      (loop escape)
                                   // compound expressions
      $|$ if mem(id) then p else p   (conditional)
      $|$ p ; p                      (sequence)
      $|$ loop p                     (repetition)
      $|$ p and p                    (par/and)
      $|$ p or p                     (par/or)
      $|$ fin p                      (finalization)
                                   // derived by semantic rules
      $|$ awaiting(id,n)             (awaiting `id' since sequence number `n')
      $|$ emitting(n)                (emitting on stack level `n')
      $|$ p @ loop p                 (unwinded loop)
\end{lstlisting}
\caption{
    Reduced syntax of \CEU.
\label{lst.formal.syntax}
}
\end{figure}

@@

The core of our semantics is a relation that, given a sequence number $n$ 
identifying the current reaction chain, maps a program $p$ and a stack of 
events $S$ in a single step to a modified program and stack:
%
\begin{align*}
\LL S, p \RR &\ST
\LL S', p' \RR
    & \textbf{(relation-inner)}
\end{align*}
%
where
%
\begin{align*}
S, S' &\in id^*
    &&(sequence~of~event~identifiers: [id_{top}, ..., id_{bottom}]) \\
p, p' &\in P
    && (program~as~described~in~Figure~\ref{lst.formal.syntax}) \\
n     &\in \mathds{N}
    && (unique~identifier~for~the~reaction~chain)
\end{align*}
%
At the beginning of a reaction chain, the stack is initialized with the 
occurring external event $ext$ ($S=[ext]$), but $emit$ expressions can push new 
events on top of it (we discuss how they are popped further).
The ever-increasing sequence number $n$ will prevent that awaiting expressions 
awake in the same reaction they are reached (ensuring that the program can 
progress).

@@

We describe this relation with a set of \emph{small-step} structural semantics 
rules, which are built in such a way that at most one transition is possible at 
any time, resulting in deterministic reaction chains.
%
The transition rules for the primary expressions are as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL S,\1await(id) \RR &\ST
\LL S,\1awaiting(id,n+1) \RR
    & \textbf{(awaiting)}      \\
%%%
\LL id:S,\1awaiting(id,m) \RR &\ST
\LL id:S,\1nop \RR, \2m<=n
    & \textbf{(awake)}   \\
%%%
\LL S,\1emit(id) \RR &\ST
\LL id:S,\1emitting(|S|) \RR
    & \textbf{(emitting)}       \\
%%%
\LL S,\1emitting(|S|) \RR &\ST
\LL S,\1nop \RR
    & \textbf{(pop)}
\end{align*}
}
%
An $await$ is simply transformed into an $awaiting$ (rule \textbf{awaiting}) as 
an artifact to remember the external sequence number $n$ it can awake.
An $awaiting$ can only transit to a $nop$ (rule \textbf{awake}) if its referred 
event $id$ matches the top of the stack and its sequence number is smaller or 
equal than the current one ($m<n$).
%
%Remember that in \CEU, the \code{await} statement returns the value associated 
%with the corresponding event: the yielded $mem$ represents the operation to 
%query that value.
%
An $emit$ transits to an $emitting$ holding the current stack level ($|S|$ 
stands for the stack length), and pushing the referred event on the stack (rule 
\textbf{emitting}).
With the new stack level $|S|+1$, the $emitting(|S|)$ itself cannot transit, as 
rule \textbf{pop} expects its parameter to match the current stack level.
This trick provides the desired stack-based semantics for internal events.

Proceeding to compound expressions, the rules for conditionals and sequences 
are straightforward:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS val(id,n) \neq 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, p \RR }
    & \textbf{(if-true)}       \\
%%%
& \frac
    { \DS val(id,n) = 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, q \RR }
    & \textbf{(if-false)}       \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~;~q) \RR \ST \LL S', (p'~;~q) \RR }
    & \textbf{(seq-adv)}      \\
%%%
& \LL S, (mem(id)~;~q) \RR \ST  \LL S, q \RR
    & \textbf{(seq-nop)}      \\
%%%
& \LL S, (break~;~q) \RR \ST \LL S, break \RR
    & \textbf{(seq-brk)}
\end{eqnarray*}
}
%
Given that our semantics focuses on control, rules \textbf{if-true} and 
\textbf{if-false} are the only to query $mem$ expressions.
%
The ``magical'' function $val$ receives the memory identifier and current 
reaction sequence number, returning the current memory value.
%
Although the value here is arbitrary, it is unique in a reaction chain, because 
a given expression can execute only once within it (remember that $loops$ must 
contain $awaits$ which, from rule \textbf{awaiting}, cannot awake in the same 
reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).
